---
title: "Growth Exploration & Initial Models"
author: "Josh Carrell - Ph.D. Student | Forest Sciences - Colorado State University"
date: 'Last Update: `r format(Sys.time(), ''%B %d, %Y'')`'
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_document:
    toc: yes
    toc_float:
      toc_collapsed: yes
    toc_depth: 4
    number_sections: no
    theme: readable
    code_folding: hide
---
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(DT)
# Data prepped for competition indexes
load("D:/Projects/SubalpineTopoGrowth/data/growth_long/subalpine_non_spatial.RData")
df <- subalpine_non_spatial
names(df)
df <- df %>% 
  filter(annual_ba_cm_2022 > 0) %>% 
  filter(dead == 0)
```


## Question & Hypotheses

1.How does soil moisture availability and competition influence subalpine tree growth?


Hypothesis 1: Tree growth will be greater in wetter sites particularly for species that are less tolerant of drought.


Hypothesis 2: Tree growth will be limited by not only soil moisture but an interaction with competition, with trees growth being suppressed in areas of low soil moisture availability and high competition.


----

## Data

```{r}
DT::datatable(df, class = 'cell-border stripe', rownames = F, filter = 'top',
              editable = T, extensions = 'Buttons', options = list(dom = 'Bfrtip',
                                                                   buttons = c("excel")))
```


## Plots

```{r, warning=FALSE}
ggplot(data = df) +
  geom_boxplot(mapping = aes(y = annual_ba_cm_2022, x = moisture_class, fill = moisture_class)) +
  theme_bw() +
  ylim(c(0,100)) +
  labs(y = "Basal Area Increase (cm^2)",
       x = "Moisture Class",
       fill = "",
       title = "2016 - 2022 Basal Area Increase ~ Moisture Class per Species") +
  facet_grid(~Spec)

ggplot(data = df) +
  geom_boxplot(mapping = aes(y = annual_ba_cm_2022, x = as.factor(soil_moisture), fill = as.factor(soil_moisture))) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust=1)) +
  ylim(c(0,100)) +
  labs(y = "Basal Area Increase (cm^2)",
       x = "Soil Moisture Level",
       fill = "",
       title = "2016 - 2022 Basal Area Increase ~ Soil Moisture per Species") +
  facet_grid(~Spec)

```

## Linear Models

```{r}
mod_lm <- lm(annual_ba_cm_2022 ~ soil_moisture + Spec, data = df)
summary(mod_lm)
```

- Basal Area increases .10 cm^2 per 1 unit increase in soil moisture level. Significant.

- Species differ in mean growth from reference (ABLA).

```{r}
mod_lm_int <- lm(annual_ba_cm_2022 ~ soil_moisture * Spec, data = df)
summary(mod_lm_int)
```

- Very small but significant influence of soil moisture on basal area increase.


```{r}
mod_lm_comp <- lm(annual_ba_cm_2022 ~ soil_moisture + ci_3 + Spec, data = df)
summary(mod_lm_comp)
```

- .10 increase in Basal Area per 1 unit increase in soil moisture.

- .2 decrease in Basal Area increase per 1 unit increase in competition.

- Species differ in significance and amount of increase when compared to reference species (ABLA).




----

## Mixed model with Plot as random effect

```{r}
mod_lmer <- lmer(annual_ba_cm_2022 ~ ci_3 + soil_moisture * Spec + dbh4 + (1 | Plot), data = df)
summary(mod_lmer)
```
This model:

- Controls for competition (ci_3)

- Allows species-specific responses to soil moisture

- Accounts for tree size (dbh4)

- Includes plot-level random intercepts (more useful when repeat measures will be included)

Results:

- Larger trees put on more basal area increase (not really useful but may help account for size)

- PIEN, PIFL, PICO positive response to soil moisture

- Competition isn't significant, most likely diminished due to plot level variations in soil moisture.


```{r}
lattice::dotplot(ranef(mod_lmer, condVar = TRUE))
```

```{r}
mod_lmer_ci_sm <- lmer(annual_ba_cm_2022 ~ ci_3 * soil_moisture + Spec + dbh4 + (1 | Plot), data = df)
summary(mod_lmer_ci_sm)
```
This model:

- Includes and interaction between competition and soil moisture.


Results:

- When competition is low, trees respond positively to soil moisture (+0.07), however when competition increases, the positive effect of moisture becomes negative. At higher competition, soil moisture has negative impact on basal area increase (moisture alone can't compensate for loss in other resources like light, space, nutrients).



```{r}
lattice::dotplot(ranef(mod_lmer_ci_sm, condVar = TRUE))
```








